Local Type Inference for Polarised System F with Existentials

### 2024-04-30 oopsla24/tex/04-algorithm.tex
Edit

In this section, we present the algorithmization of the declarative system described above.
The algorithmic system follows the structure of the declarative specification closely.
First, it is also given by a set of inference rules, which, however,
must be mode-correct (\cite{dunfield2020:bidirectional}), \ie
the output of each rule is always uniquely defined by its input.
And second, most of the declarative rules 
(except for the rules \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel})
have a unique algorithmic counterpart, 
which simplifies reasoning about the algorithm and its correctness proofs.

@@

\begin{figure}[h]
  \begin{multicols}{2}
      Negative Algorithmic Variables\\
      $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots\\
      \columnbreak
      Positive Algorithmic Variables\\
      $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots\\
  \end{multicols}
  \hfill\\
  \begin{multicols}{2}
      \ottuNShort
      \columnbreak
      \ottuPShort
  \end{multicols}
  \hfill\\
  Algorithmic Type Context\\
   $[[Ξ]] = \{[[α1̂±]], \dots, [[αn̂±]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \hfill\\
  Constraint Type Context\\
   $[[Θ]] \Coloneqq \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \caption{Algorithmic Syntax}
  \label{fig:algo-syntax}
\end{figure}

@@

\paragraph{Algorithmic Contexts and Well-formedness}
To specify when algorithmic types are well-formed, 
we define algorithmic contexts $[[Ξ]]$ as sets of algorithmic variables.
Then $[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness
judgment of algorithmic terms defined as expected. 
Informally, they check that all free declarative variables are in $[[Γ]]$, 
and all free algorithmic variables are in $[[Ξ]]$.
Most of the rules are inherited from the well-formedness of 
\emph{declarative} types: the declarative variables are
checked to belong to the context $[[Γ]]$, the quantifiers extend the context,
type constructors are well-formed congruently.
As we extend the syntax with algorithmic variables, we also add 
two base-case rules for them:
 \ruleref{\ottdruleWFATPUVarLabel} and \ruleref{\ottdruleWFATNUVarLabel} 
 (see \cref{fig:algo-wf}).

\begin{figure}[h]
\begin{multicols}{2}
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleWFATPUVarLabeled{}}
  \columnbreak\\
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleWFATNUVarLabeled{}}
\end{multicols}
\caption{Well-formedness of Algorithmic Types}
\label{fig:algo-wf}
\end{figure}

\paragraph{Algorithmic Normalization}
Similarly to well-formedness, the normalization of algorithmic types is defined
by extending the declarative definition with the algorithmic variables.
To the rules repeating the declarative normalization, we add
rules saying that normalization is trivial on algorithmic variables (see \cref{fig:algo-nf}).

\begin{figure}[h]
\begin{multicols}{2}
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleNrmPUVar{}}
  \columnbreak\\
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleNrmNUVar{}}
\end{multicols}
\label{fig:algo-nf}
\caption{Normalization of Algorithmic Types}
\end{figure}

\subsection{Type Constraints}
As the algorithm proceeds, it accumulates the information 
about the algorithmic type variables in the form of \emph{constraints}.
In our system, the constraints can be of two kinds: 
\emph{subtyping constraints} and \emph{unification constraints}.
The subtyping constraint can only have a positive shape $[[α̂⁺ :≥ iP]]$, \ie it 
restricts a positive algorithmic variable to be a 
supertype of a certain declarative type---this is one of the invariants that we 
preserve in the algorithm.
The unification constraint can have either a positive form $[[α̂⁺ :≈ iP]]$ or
a negative form $[[α̂⁻ :≈ iN]]$, however, the right-hand side of the constraint
cannot contain algorithmic type variables.
The set of constraints is denoted as $[[SC]]$. 
We assume that each algorithmic variable can be restricted by at most one constraint.

We separately define $[[UC]]$ as a set consisting of unification constraints only.
This is done to simplify the representation of the algorithm. 
The unification algorithm, which we use as a subroutine of the subtyping algorithm,
can only produce unification constraints.
A set of unification constraints can be resolved in a simpler way than
a general constraint set. This way, 
the separation of the unification constraint resolution into a separate
procedure allows us to better decompose the structure of the algorithm,
and thus, simplify the inductive proofs.

@@

\paragraph{Constraint Contexts}
When one instantiates an algorithmic variable, 
they may only use type variables available in its scope.
As such, each algorithmic variable must remember the context at the moment when 
it was introduced. In our algorithm, this information is represented by
a \emph{constraint context} $[[Θ]]$---a set of pairs associating 
algorithmic variables and declarative contexts.

\paragraph{Auxiliary Functions}
We define $[[dom(SC)]]$---a domain of a constraint set $[[SC]]$ as a set of algorithmic variables
that it restricts. Similarly, we define $[[dom(Θ)]]$---a domain of constraint context
as a set of algorithmic variables that $[[Θ]]$ associates with their contexts.
We write $[[Θ(α̂±)]]$ to denote the context associated with $[[α̂±]]$ in $[[Θ]]$.

@@

  For convenience and scalability, 
  we decompose the subtyping algorithm 
  into several procedures. \Cref{fig:alg-subtyping-graph}
  shows these procedures and the dependencies between them:
  arrows denote the invocation of one procedure from another.
  The label <<$\ottkw{nf}$>> annotating arrows means that the calling
  procedure normalizes the input before passing it to the callee.

@@

As an input, the subtyping algorithm takes
a type context $[[Γ]]$, a constraint context $[[Θ]]$,
and two types of the corresponding polarity:
$[[uN]]$ and  $[[iM]]$ for the negative subtyping, and
$[[uP]]$ and  $[[iQ]]$ for the positive subtyping.
We assume the second type ($[[iM]]$ and $[[iQ]]$) to be 
declarative (with no algorithmic variables) and well-formed in $[[Γ]]$,
but the first type ($[[uN]]$ and $[[uP]]$) may contain algorithmic variables,
whose instantiation contexts are specified by $[[Θ]]$.

@@

\paragraph{Quantifiers}  
Rules \ruleref{\ottdruleAForallLabel} and 
\ruleref{\ottdruleAExistsLabel} are symmetric. 
Declaratively, the quantified variables on the left-hand side must 
be instantiated with types, which are not known beforehand.
We address this problem by algorithmization 
of the quantified variables (see \cref{sec:algo-syntax}).
The rule introduces fresh algorithmic variables
$[[puas]]$ or $[[nuas]]$,
puts them into the constraint context $[[Θ]]$
(specifying that they must be instantiated in the extended context
$[[Γ, pbs]]$ or $[[Γ, nbs]]$) and substitute the quantified variables
for them in the input type.

After algorithmization of the quantified variables, 
the algorithm proceeds with the recursive call, returning
constraints $[[SC]]$. As the output, the algorithm removes the freshly
introduced algorithmic variables from the constraint context, This operation is
sound: it is guaranteed that $[[SC]]$ always has a solution, but the specific
instantiation of the freshly introduced algorithmic variables is not important,
as they do not occur in the input types.

\paragraph{Functions}
To infer the subtyping of the function types, the algorithm
makes two calls: 
\begin{enumerate*}
  \item[(i)] a recursive call ensuring the subtyping of the result types, and
  \item[(ii)] a call to positive subtyping (or rather super-typing) on the argument types.
\end{enumerate*}
The resulting constraints are merged
(using a special procedure defined in \cref{sec:constraint-merge})
and returned as the output.

\paragraph{Algorithmic Variable}
If one of the sides of the subtyping is a unification variable,
the algorithm must create a new constraint. 
Because the right-hand side of the subtyping is always declarative,
it is only the left-hand side that can be a unification variable.
Moreover, another invariant we preserve prevents the negative
algorithmic variables from occurring in types during the
 negative subtyping algorithm. It means
that the only possible form of the subtyping here is $[[α̂⁺]] [[≥]] [[iP]]$,
which is covered by \ruleref{\ottdruleAPUVarLabel}.

The potential problem here is that the type $[[iP]]$
might be not well-formed in the context required for $[[α̂⁺]]$ by $[[Θ]]$, 
because this context might be smaller than the current context $[[Γ]]$.
As we wish the resulting constraint set to be sound \wrt $[[Θ]]$,
we cannot simply put $[[α̂⁺ :≥ iP]]$ into the output. 
Prior to that, we update the type $[[iP]]$ to its lowest supertype $[[iQ]]$
well-formed in $[[Θ(α̂⁺)]]$. It is done by the \emph{upgrade} procedure,
which we discuss in detail in \cref{sec:lub}.

@@

As an input the unification context 
takes a type context $[[Γ]]$, a constraint context $[[Θ]]$,
and two types of the required polarity:
$[[uN]]$ and  $[[iM]]$ for the negative unification, and
$[[uP]]$ and  $[[iQ]]$ for the positive unification.
It is assumed that only the left-hand side type may contain algorithmic variables,
this way, the left-hand side is well-formed as an algorithmic type in $[[Γ]]$ and 
$[[Θ]]$, whereas the right-hand side is well-formed declaratively in $[[Γ]]$.

Since only the left-hand side may contain algorithmic variables,
that the unification instantiates, we could have called this procedure \emph{matching}.
However, in \cref{sec:weakening-invariant}, we will discuss several modifications of the 
type system, where this invariant is not preserved, and therefore, this procedure
requires general first-order pattern unification \cite{miller1991:unification}.

As the output, the unification algorithm returns the weakest
set of unification constraints $[[UC]]$ such that any instantiation 
satisfying these constraints unifies the input types.

@@

\paragraph{Shifts}
  The shift rules \ruleref{\ottdruleUShiftDLabel} and \ruleref{\ottdruleUShiftULabel}
  require the input types to be formed by the same shift constructor. 
  They remove this constructor, unify the types recursively, and return the resulting
  set of constraints.

@@

\paragraph{Functions}
  Rule \ruleref{\ottdruleUArrowLabel} unifies two functional types. 
  First, it unifies the argument types and their result types recursively.
  Then it merges the resulting constraints using the procedure described in \cref{sec:constraint-merge}.

  Notice that the resulting constraints can only have \emph{unification} entries.
  It means that they can be merged in a simpler way than general constraints.
  In particular, the merging procedure does not call any of the subtyping subroutines,
  but rather simply checks the matching constraint entries for equality.

\paragraph{Algorithmic Variable}
  Finally, if the left-hand side of the unification is an algorithmic variable,
  \ruleref{\ottdruleUNVarLabel} or \ruleref{\ottdruleUPVarLabel} is applied. 
  It simply checks that the right-hand side type is well-formed in the required
  constraint context, and returns a newly created constraint restricting the variable
  to be equal to the right-hand side type.

\vspace{\baselineskip}
As one can see, the unification procedure is standard, 
except that it makes sure that the resulting instantiations agree with the input
constraint context $[[Θ]]$.  As a subroutine, the unification algorithm only uses the (unification) 
constraint merge procedure and the well-formedness checking.

@@

In this section, we discuss the constraint merging procedure.
It allows one to combine two constraint sets into one. 
A simple union of two constraint sets is not sufficient, 
since the resulting set must not contain two entries restricting the 
same algorithmic variable---we call such entries \emph{matching}.
The matching entries must be combined into \emph{one} constraint entry, that 
would represent their conjunction. 
This way, to merge two constraint sets, we unite the entries of two
sets, and then merge the matching pairs.

\paragraph{Merging Matching Constraint Entreies}
Two \emph{matching} entries formed in the same context $[[Γ]]$ 
can be merged as shown in \cref{fig:merge-entries}.
Suppose that $[[scE1]]$ and $[[scE2]]$ are input entries. 
The result of the merge $[[scE1 & scE2]]$ must be the 
weakest entry which implies both  $[[scE1]]$ and $[[scE2]]$.

@@

Suppose that one of the input entries, say $[[scE1]]$, is a unification 
constraint entry. Then the resulting entry $[[scE1]]$ must coincide with it 
(up-to-equivalence), and thus, it is only required to check that $[[scE2]]$ 
is implied by $[[scE1]]$.
\begin{itemize}
  \item If $[[scE2]]$ is also a restricting entry, then the types on the right-hand side
    of $[[scE1]]$ and $[[scE2]]$ must be equivalent,
    as given by rules \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}.
  \item If $[[scE2]]$ is a supertype constraint $[[α̂⁺ :≥ iP]]$,
    the algorithm must check that the type assigned by $[[scE1]]$ is a supertype of $[[iP]]$.
    The corresponding symmetric rules are \ruleref{\ottdruleSCMESupEqLabel} and \ruleref{\ottdruleSCMEEqSupLabel}.
\end{itemize}

If both input entries are supertype constraints: $[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$,
then their conjunction is $[[α̂⁺ :≥ iP ∨ iQ]]$, as given by \ruleref{\ottdruleSCMESupSupLabel}.
The least upper bound---$[[iP ∨ iQ]]$ is the least supertype of both $[[iP]]$ and $[[iQ]]$,
and this way, $[[α̂⁺ :≥ iP ∨ iQ]]$ is the weakest constraint entry that implies
$[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$. The algorithm finding the least upper bound
is discussed in \cref{sec:lub}.

\paragraph{Merging Constraint Sets}
  The algorithm for merging constraint sets is shown in \cref{fig:merge-subtyping-constraints}.
  As discussed, the result of merge $[[SC1]]$ and $[[SC2]]$ consists of three parts: 
  \begin{enumerate*}
    \item[(i)] the entries of $[[SC1]]$ that do not match any entry of $[[SC2]]$;
    \item[(ii)] the entries of $[[SC2]]$ that do not match any entry of $[[SC1]]$; and
    \item[(iii)] the merge (\cref{fig:merge-entries}) of matching entries.
  \end{enumerate*}

@@

As shown in \cref{fig:merge-entries}, the merging procedure relies 
substantially on the least upper bound algorithm.
In the next section, we discuss this algorithm in detail,
together with the upgrade procedure, selecting the least supertype 
ell-formed in a given context.

@@

The application type inference judgment has form 
$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$.
As an input, it takes three contexts: typing context $[[Γ]]$, a variable context $[[Φ]]$,
and a constraint context $[[Θ1]]$. It also takes a head type $[[uN]]$ and 
a list of arguments (terms) $[[args]]$ the head is applied to.
The head may contain algorithmic variables specified by $[[Θ1]]$, 
in other words, $[[Γ; dom(Θ1) ⊢ uN]]$.
As a result, the application inference judgment returns 
$[[uM]]$---a normalized type of the result of the application.
Type $[[uM]]$ may contain new algorithmic variables, and thus, 
the judgment also returns $[[Θ2]]$---an updated constraint context
and $[[SC]]$---a set of subtyping constraints.
Together $[[Θ2]]$  and $[[SC]]$ specify how the algorithmic variables 
must be instantiated.

>>>>

In this section, we present the algorithmization of the declarative system described above.
The algorithmic system follows the structure of the declarative specification closely.
First, it is also given by a set of inference rules, which, however,
are mode-correct (\cite{dunfield2020:bidirectional}), \ie
the output of each rule is always uniquely defined by its input.
And second, most of the declarative rules 
(except for the rules \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel})
have a unique algorithmic counterpart, 
which simplifies reasoning about the algorithm and its correctness proofs.

@@

\begin{figure}
  \begin{multicols}{2}
      Negative Algorithmic Variables\\
      $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots\\
      \columnbreak
      Positive Algorithmic Variables\\
      $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots\\
  \end{multicols}
  \hfill\\
  \begin{multicols}{2}
      \ottuNShort
      \columnbreak
      \ottuPShort
  \end{multicols}
  \hfill\\
  Algorithmic Type Context\\
   $[[Ξ]] \Coloneqq \{[[α1̂±]], \dots, [[αn̂±]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \hfill\\
  Instantiation Context\\
   $[[Θ]] \Coloneqq \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
  \caption{Algorithmic Syntax}
  \label{fig:algo-syntax}
\end{figure}

@@

\paragraph{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
To specify when algorithmic types are well-formed, we define algorithmic
contexts $[[Ξ]]$ as sets of algorithmic variables. Then
$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness judgment of
algorithmic terms defined as expected. Informally, they check that all free
declarative variables are in $[[Γ]]$, and all free algorithmic variables are in
$[[Ξ]]$. Most of the rules are inherited from the well-formedness of
\emph{declarative} types: the declarative variables are checked to belong to the
context $[[Γ]]$, the quantifiers extend the context, type constructors are
well-formed congruently. As we extend the syntax with algorithmic variables, we
also add two base-case rules for them: \ruleref{\ottdruleWFATPUVarLabel} and
\ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf}).

\begin{figure}
\begin{multicols}{2}
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleWFATPUVarLabeled{}}
  \columnbreak\\
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleWFATNUVarLabeled{}}
\end{multicols}
\caption{Well-formedness of Algorithmic Types}
\label{fig:algo-wf}
\end{figure}

\paragraph{Instantiation Context $[[Θ]]$}
When one instantiates an algorithmic variable, one may only use type variables
\emph{available in its scope}. As such, each algorithmic variable must remember
the context at the moment when it was introduced. In our algorithm, this
information is represented by an \emph{instantiation context} $[[Θ]]$---a set of
pairs associating algorithmic variables and declarative contexts.

\paragraph{Algorithmic Normalization}
Similarly to well-formedness, the normalization of algorithmic types is defined
by extending the declarative definition (\cref{fig:type-nf}) with the
algorithmic variables. To the rules repeating the declarative normalization, we
add rules saying that normalization is trivial on algorithmic variables.

\begin{figure}
\begin{multicols}{2}
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleNrmPUVar{}}
  \columnbreak\\
  $\vcenter{\hbox{\vdots}}$\\
  \ottusedrule{\ottdruleNrmNUVar{}}
\end{multicols}
\caption{Normalization of Algorithmic Types}
\label{fig:algo-nf}
\end{figure}

\subsection{Type Constraints}
Throughout the algorithm's operation, it gathers information about the
algorithmic type variables, represented as \emph{constraints}. These constraints
in our system can be either \emph{subtyping constraints} or \emph{unification
constraints}. As preserved by the algorithm, each subtyping constraint has a
positive shape $[[α̂⁺ :≥ iP]]$, meaning it confines a positive algorithmic
variable to be the supertype of a positive declarative type. Unification
constraints can take a positive $[[α̂⁺ :≈ iP]]$ or negative $[[α̂⁻ :≈ iN]]$
form, but algorithmic type variables cannot occur on their right-hand side. The
constraints \emph{set} is denoted as $[[SC]]$, and we presume that each
algorithmic variable can be restricted by at most one constraint.

We define $[[UC]]$ separately as a set solely containing \emph{unification}
constraints for simpler algorithm representation. The unification algorithm,
which we use as a subroutine of the subtyping algorithm, can only produce
unification constraints. A resolution of unification constraints is simpler than
that of a general constraint set. This way, this separation allows us to better
decompose the algorithm's structure, thus simplifying the inductive proofs.

@@

\paragraph{Auxiliary Functions}
We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of algorithmic variables
that it restricts. Similarly, we define $[[dom(Θ)]]$---the domain of instantiation context
---as a set of algorithmic variables that $[[Θ]]$ associates with their contexts.
We write $[[Θ(α̂±)]]$ to denote the declarative context associated with $[[α̂±]]$ in $[[Θ]]$.

@@

  For convenience and scalability, 
  we decompose the subtyping algorithm 
  into several procedures. \Cref{fig:alg-subtyping-graph}
  shows these procedures and the dependencies between them:
  arrows denote the invocation of one procedure from another.

  Some of the procedures (in particular, the unification and the
  anti-unification) assume that the input types are normalized. Therefore, we
  call the normalization procedure before invoking them, indicating this by the
  `$\ottkw{nf}$' annotation on the arrows in \cref{fig:alg-subtyping-graph}.
  Alternatively, one could normalize the input types in the very beginning,
  as the normalization is preserved throughout the algorithm. However, we
  delay the normalization to the places where it is required to show that
  normalization is needed only at these stages to maintain consistent
  invariants.

@@

As an input, the subtyping algorithm takes
a type context $[[Γ]]$, an instantiation context $[[Θ]]$,
and two types of the corresponding polarity:
$[[uN]]$ and  $[[iM]]$ for the negative subtyping, and
$[[uP]]$ and  $[[iQ]]$ for the positive subtyping.
We assume the second type ($[[iM]]$ and $[[iQ]]$) to be 
declarative (with no algorithmic variables) and well-formed in $[[Γ]]$,
but the first type ($[[uN]]$ and $[[uP]]$) may contain algorithmic variables,
whose instantiation contexts are specified by $[[Θ]]$.

@@

\paragraph{Quantifiers}  
Rules \ruleref{\ottdruleAForallLabel} and 
\ruleref{\ottdruleAExistsLabel} are symmetric. 
Declaratively, the quantified variables on the left-hand side must 
be instantiated with types, which are not known beforehand.
We address this problem by algorithmization 
of the quantified variables (see \cref{sec:algo-syntax}).
The rule introduces fresh algorithmic variables
$[[puas]]$ or $[[nuas]]$,
puts them into the instantiation context $[[Θ]]$
(specifying that they must be instantiated in the extended context
$[[Γ, pbs]]$ or $[[Γ, nbs]]$) and substitute the quantified variables
for them in the input type.

After algorithmization of the quantified variables, 
the algorithm proceeds with the recursive call, returning
constraints $[[SC]]$. As the output, the algorithm removes the freshly
introduced algorithmic variables from the instantiation context, This operation is
sound: it is guaranteed that $[[SC]]$ always has a solution, but the specific
instantiation of the freshly introduced algorithmic variables is not important,
as they do not occur in the input types.

\paragraph{Functions}
To infer the subtyping of the function types, the algorithm
makes two calls: 
\begin{enumerate*}
  \item[(i)] a recursive call ensuring the subtyping of the result types, and
  \item[(ii)] a call to positive subtyping (or rather super-typing) on the argument types.
\end{enumerate*}
The resulting constraints are merged
using a special procedure defined in \cref{sec:constraint-merge}
and returned as the output.

\paragraph{Algorithmic variables}
If one of the sides of the subtyping is a unification variable, the algorithm
creates a new constraint. Because the right-hand side of the subtyping is always
declarative, it is only the left-hand side that can be a unification variable.
Moreover, another invariant we preserve prevents the negative algorithmic
variables from occurring in types during the negative subtyping algorithm. It
means that the only possible form of the subtyping here is $[[α̂⁺]] [[≥]]
[[iP]]$, which is covered by \ruleref{\ottdruleAPUVarLabel}.

The potential problem here is that the type $[[iP]]$ might be not well-formed in
the instantiation context required for $[[α̂⁺]]$ by $[[Θ]]$ because this
context might be smaller than the current context $[[Γ]]$. As we wish the
resulting constraint set to be sound \wrt $[[Θ]]$, we cannot simply put 
$[[α̂⁺ :≥ iP]]$ into the output. Prior to that, we update the type $[[iP]]$ to its
lowest supertype $[[iQ]]$ well-formed in $[[Θ(α̂⁺)]]$. It is done by the
\emph{upgrade} procedure, which we discuss in detail in \cref{sec:lub}.

@@

As an input, the unification context takes a type context $[[Γ]]$, an
instantiation context $[[Θ]]$, and two types of the required polarity: $[[uN]]$
and  $[[iM]]$ for the negative unification, and $[[uP]]$ and  $[[iQ]]$ for the
positive unification. It is assumed that only the left-hand side type may
contain algorithmic variables. This way, the left-hand side is well-formed as an
algorithmic type in $[[Γ]]$ and $[[Θ]]$, whereas the right-hand side is
well-formed declaratively in $[[Γ]]$.

Since only the left-hand side may contain algorithmic variables that the unification instantiates, we could have called this procedure \emph{matching}.
However, in \cref{sec:weakening-invariant}, we will discuss several
modifications of the type system, where this invariant is not preserved, and
therefore, this procedure requires general first-order pattern unification
\cite{miller1991:unification}.

As the output, the unification algorithm returns \emph{the weakest} set of
unification constraints $[[UC]]$ such that \emph{any} instantiation satisfying
these constraints unifies the input types.

@@

\paragraph{Shifts}
  The shift rules \ruleref{\ottdruleUShiftDLabel} and \ruleref{\ottdruleUShiftULabel}
  require the two input types to be formed by the same shift constructor. 
  They remove this constructor, unify the types recursively, and return the resulting
  set of constraints.

@@

\paragraph{Functions}
  Rule \ruleref{\ottdruleUArrowLabel} unifies two functional types. 
  First, it unifies their argument types and their result types recursively.
  Then it merges the resulting constraints using the procedure described in \cref{sec:constraint-merge}.

  Notice that the resulting constraints can only have \emph{unification}
  entries. It means that they can be merged in a simpler way than general
  constraints. In particular, the merging procedure does not call any of the
  subtyping subroutines but rather simply checks the matching constraint entries
  for equality.

\paragraph{Algorithmic variable}
  Finally, if the left-hand side of the unification is an algorithmic variable,
  \ruleref{\ottdruleUNVarLabel} or \ruleref{\ottdruleUPVarLabel} is applied. 
  It simply checks that the right-hand side type is well-formed in the required
  instantiation context, and returns a newly created constraint restricting the variable
  to be equal to the right-hand side type.

\vspace{\baselineskip}
As one can see, the unification procedure is standard; the only peculiarity
(although it is common for type inference) is that it makes sure that the resulting
instantiations agree with the input instantiation context $[[Θ]]$.  As a
subroutine, the unification algorithm only uses the (unification) constraint
merge procedure and the well-formedness checking.

@@

In this section, we discuss the constraint merging procedure.
It allows one to combine two constraint sets into one. 
A simple union of two constraint sets is not sufficient, 
since the resulting set must not contain two entries restricting the 
same algorithmic variable---we call such entries \emph{matching}.

The matching entries must be combined into \emph{one} constraint entry, that
would represent their conjunction. This way, to merge two constraint sets, we
unite the entries of two sets and then merge the matching pairs.

\paragraph{Merging matching constraint entries}
Two \emph{matching} entries formed in the same context $[[Γ]]$ 
can be merged as shown in \cref{fig:merge-entries}.
Suppose that $[[scE1]]$ and $[[scE2]]$ are input entries. 
The result of the merge $[[scE1 & scE2]]$ is 
\emph{the weakest entry which implies both $[[scE1]]$ and $[[scE2]]$}.

@@

Suppose that one of the input entries, say $[[scE1]]$, is a \emph{unification}
constraint entry. Then the resulting entry $[[scE1]]$ must coincide with it 
(up-to-equivalence), and thus, it is only required to check that $[[scE2]]$ 
is implied by $[[scE1]]$. We consider two options:
\begin{itemize}
  \item[(i)] if $[[scE2]]$ is also a \emph{unification} entry, then the types on
    the right-hand side of $[[scE1]]$ and $[[scE2]]$ must be equivalent, as
    given by rules \ruleref{\ottdruleSCMEPEqEqLabel} and
    \ruleref{\ottdruleSCMENEqEqLabel};
  \item[(ii)] if $[[scE2]]$ is a \emph{supertype} constraint entry $[[α̂⁺ :≥ iP]]$,
    the algorithm must check that the type assigned by $[[scE1]]$ is a supertype of $[[iP]]$.
    The corresponding symmetric rules are \ruleref{\ottdruleSCMESupEqLabel} and \ruleref{\ottdruleSCMEEqSupLabel}.
    In the premises of these rules, $[[uP]]$ is the same as $[[iP]]$ below, 
    and $[[uQ]]$ is the same as $[[iQ]]$ below but relaxed to an algorithmic type.
\end{itemize}

If both input entries are supertype constraints: $[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$,
then their conjunction is $[[α̂⁺ :≥ iP ∨ iQ]]$, as given by \ruleref{\ottdruleSCMESupSupLabel}.
The least upper bound---$[[iP ∨ iQ]]$---is the least supertype of both $[[iP]]$ and $[[iQ]]$,
and this way, $[[α̂⁺ :≥ iP ∨ iQ]]$ is the weakest constraint entry that implies
$[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$. The algorithm finding the least upper bound
is discussed in \cref{sec:lub}.

\paragraph{Merging constraint sets}
  The algorithm for merging constraint sets is shown in \cref{fig:merge-subtyping-constraints}.
  As discussed, the result of merge $[[SC1]]$ and $[[SC2]]$ consists of three parts: 
  \begin{enumerate*}
    \item[(i)] the entries of $[[SC1]]$ that do not match any entry of $[[SC2]]$;
    \item[(ii)] the entries of $[[SC2]]$ that do not match any entry of $[[SC1]]$; and
    \item[(iii)] the merge (\cref{fig:merge-entries}) of matching entries.
  \end{enumerate*}

@@

As shown in \cref{fig:merge-entries}, the merging procedure relies 
substantially on the least upper bound algorithm.
In the next section, we discuss this algorithm in detail,
together with the upgrade procedure, selecting the least supertype 
well-formed \emph{in a given context}.

@@

The application type inference judgment has form 
$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$.
As an input, it takes three contexts: typing context $[[Γ]]$, a variable context $[[Φ]]$,
and an instantiation context $[[Θ1]]$. It also takes a head type $[[uN]]$ and 
a list of arguments (terms) $[[args]]$ the head is applied to.
The head may contain algorithmic variables specified by $[[Θ1]]$, 
in other words, $[[Γ; dom(Θ1) ⊢ uN]]$.
As a result, the application inference judgment returns 
$[[uM]]$---a normalized type of the result of the application.
Type $[[uM]]$ may contain new algorithmic variables, and thus, 
the judgment also returns $[[Θ2]]$---an updated instantiation context
and $[[SC]]$---a set of subtyping constraints.
Together $[[Θ2]]$  and $[[SC]]$ specify how the algorithmic variables 
must be instantiated.

### 2024-04-30 oopsla24/tex/05-proof.tex
Edit

The central results ensuring the correctness of the inference algorithm are its soundness and completeness with respect to the declarative specification.
The soundness means the algorithm will always
produce a typing \emph{allowed} by the declarative system;
Dually, the completeness says that once a term has some type declaratively, the inference algorithm succeeds.

>>>>

\label{sec:proofs}
The central results ensuring the correctness of the inference algorithm are its soundness and completeness with respect to the declarative specification.
The soundness means the algorithm will always
produce a typing \emph{allowed} by the declarative system;
Dually, the completeness says that once a term has some type declaratively, the inference algorithm succeeds.
